A Discontinuous Petrov-Galerkin method for compressible flow problems

### 2012-08-14 docs/Jesse/Proposal/boeingPres.tex
adding boeing presentation to git

\title[A DPG method for Compressible Flow]{A Discontinuous Petrov-Galerkin method for compressible flow problems}
\date{August 8, 2012}
\author[J. Chan, L. Demkowicz, R. Moser]{Jesse Chan \\ Supervisors: Leszek Demkowicz, Robert Moser \\ Committee: Todd Arbogast, Omar Ghattas, Venkat Raman}

@@

}
\frame{
\frametitle{Compressible Navier-Stokes equations}
\begin{columns}[c]
\begin{column}{.475\textwidth}
Numerical difficulties:
\begin{itemize}
\item{} Resolving solution features (sharp, localized viscous-scale phenomena)
\begin{itemize}
\item{} Shocks
\item{} Boundary layers - resolution needed for drag/load
\item{} Turbulence (non-localized)
\end{itemize}
\item{} Nonlinear convergence and uniqueness of solutions
\item{} Stability of numerical schemes
\begin{itemize}
\item{} Coarse/adaptive grids
\item{} Higher order
\end{itemize}
\end{itemize}
\end{column}
\begin{column}{.475\textwidth}
\begin{figure}
\centering
\includegraphics[scale = .7]{figs/compFlowPics/airfoil_mesh.png}
\end{figure}
\end{column}
\end{columns}
\vspace{.2cm}
\begin{center}
Idea: begin first with the model problem of convection-diffusion.
\end{center}
}
\frame{
\frametitle{Robustness: convection-diffusion as a model problem}
\vspace{-.5cm}
\[
\div \left(\beta u\right) - \epsilon \Delta u = f, \quad \text{on }\Omega \in \mathbb{R}^3
\]
\vspace{-.5cm}
\begin{columns}[c]
\begin{column}{.49\textwidth}
In 1D: $\beta u' - \epsilon u'' = f$. Standard continuous Galerkin variational formulation: solve
\[
b(u,v) = \ell(v), \quad u,v\in H^1_0([0,1])
\]
where
\begin{align*}
b(u,v) &= \int_\Omega -\beta uv' + \epsilon u'v' \\
\ell(v) &= \int_\Omega f v 
\end{align*}
\end{column}
\begin{column}{.49\textwidth}
\begin{figure}
\centering
\includegraphics[scale=.35]{figs/GalerkinOscTight.png}
\caption{Solution for $f=1$. Oscillations in the standard Galerkin method for large Peclet numbers ${\rm Pe} \coloneqq \frac{h}{\epsilon}$.}
\end{figure}
\end{column}
\end{columns}
}
%\frame{
%\frametitle{Stabilization}
%Historical stabilization methods: 
%\begin{itemize}
%\item Artificial diffusion - solve
%\[
%\beta u' - \tilde{\epsilon} u'' = f
%\]
%where $\tilde{\epsilon}$ is set depending on $\beta$, $\epsilon$, and the mesh. Can be made ``exact'' for $f=0$. 
%\item Upwind finite differences - for $\beta > 0$, and uniform grid points $x_A$,
%\[
%u'(x_A) \approx \frac{u_h(x_A)-u_h(x_{A-1})}{h}
%\]
%\end{itemize}
%Both of these methods introduce additional numerical diffusion, and their effectiveness depends on the forcing $f$ and parameters $h$, $|\beta|$, and $\epsilon$.
%}
\frame{
\frametitle{Streamline-upwind Petrov-Galerkin (SUPG)}
SUPG solves $b_{\rm SUPG}(u,v) = l_{\rm SUPG}(v)$, where
\begin{align*}
b_{\rm SUPG}(u,v) &= b(u,v)+ \sum_{K} \int_{K} \tau \LRp{L_{\rm adv} v} L u \\
l_{\rm SUPG}(v) &= \ell(v) + \sum_{K} \int_{K} \tau \LRp{L_{\rm adv}v} f.
\end{align*}
\vspace{-.5cm}
\begin{columns}[c]
\begin{column}{.44\textwidth}
\vspace{-.5cm}
\begin{itemize}
\item{} $Lu = \div\left(\beta u\right) - \epsilon \triangle u$, $L_{\rm adv}u = \div \left(\beta u\right)$, and $\tau$ is a parameter. 
%\item{} Recovers ``exact'' artificial diffusion for $f=0$.
\item{} Effective for $f\neq 0$. %``exact'' artificial diffusion.
\item{} \emph{Residual-based} stabilization. 
\end{itemize}
\end{column}
\begin{column}{.54\textwidth}
\begin{figure}
\centering
\includegraphics[scale=.38]{figs/SUPG.png}
%\caption{Exact, Bubnov-Galerkin, and SUPG solutions.}
\end{figure}
\end{column}
\end{columns}
}

@@

%\frame{
%\frametitle{Regularization}
%Discontinuous inflow conditions on $\widehat{f_n}$, out of alignment with mesh. Boundary condition $\epsilon \grad u \cdot n = 0$ at outflow.
%\begin{figure}
%\centering
%\subfigure{
%\includegraphics[scale=.33]{figs/discontinuous.png}
%}
%\subfigure{
%\includegraphics[scale=.305]{figs/discontinuousL2rates.png}
%}
%\caption{Discontinuous hat inflow data with $\epsilon = 1e-5$, convection solution errors.}
%\end{figure}
%}

@@

\begin{figure}
\centering
\subfigure{
\includegraphics[scale=.21]{figs/vortex.png}
}
\subfigure{
\includegraphics[scale=.2015]{figs/vortexMesh.png}
}
\caption{Steady vortex problem with $\epsilon = 1e-4$, using a variant of the test norm.}
\end{figure}
\vspace{-.25cm}
}

@@

\section{Proposed work}
\subsection{Area A}
\frame{
\frametitle{Proposed work: Area A}
\begin{itemize}
\item{\textbf{Completed: Prove robustness of DPG method for the scalar convection-diffusion problem.}}

We have introduced a test norm under which the DPG method robustly bounds the $L^2$ error in the field variables $u$ and the scaled stress $\sigma$. Numerical results confirm the theoretical bounds given.

\item{\textbf{\textcolor{red}{Proposed}: Attempt analysis of the linearized Navier-Stokes system.}}

We hope to analyze the linearized Navier-Stokes equations to determine an optimal extrapolation of the test norm for the scalar convection-diffusion problem to systems.

\end{itemize}
}

\subsection{Area B}

\frame{
\frametitle{Proposed work: Area B}
\begin{itemize}
\item{\textbf{Completed: Collaborative work with Nathan Roberts on the higher order parallel adaptive DPG code Camellia.}}

Numerical experiments use the higher-order adaptive codebase Camellia, built by Nathan Roberts upon the Trilinos library. %The framework for arbitrary-irregularity anisotropic refinements in both $h$ and $p$ is in place, and the code is partially parallelized.

\item{\textbf{\textcolor{red}{Proposed}: Anisotropic refinements and $hp$-schemes.}}

The error representation function drives refinement effectively, and we hope to generalize its use to anisotropic adaptive schemes. 
\begin{figure}
\centering
\subfigure{\includegraphics[scale = .175]{figs/anisotropy/b5.png}}
\subfigure{\includegraphics[scale = .18]{figs/anisotropy/v6.png}}
\caption{Anisotropic mesh for a convection-diffusion boundary layer.}
\end{figure}
\end{itemize}
}

\frame{
\textbf{\textcolor{red}{Proposed}: Distributed iterative static condensation.}
\[
Ku = \arr{A}{B}{B^T}{D}\vecttwo{u_{\rm flux}}{u_{\rm field}} = \vecttwo{f}{g} = l
\]
where $D$ has a block-diagonal structure. The system can be reduced to yield the condensed system
\[
\left(A-B D^{-1} B^T\right)u_{\rm flux} = f - B D^{-1} g
\]
where $D^{-1}$ can be inverted block-wise. For FE stiffness matrices under the Laplace equation, the Schur complement has reduced condition number of $O(h^{-1})$ as opposed to $O(h^{-2})$.\footnote{\bibentry{schurComplement}}
}

\frame{  
\textbf{\textcolor{red}{Proposed}: a Nonlinear Hessian-based DPG method.}
Given a nonlinear variational problem $b(u,v) = \ell(v)$, linear in $v$ but not in $u$, beginning with the \textit{nonlinear} dual residual 
\[
J(u_h) = \frac{1}{2}\|B(u_h)-\ell\|_{V'}^2 \coloneqq\frac{1}{2} \sup_{v\in V\setminus\{0\}} \frac{| b(u_h,v)-\ell(v)|^2}{\nor{v}_V^2}.
\]
produces a Hessian-based DPG method, which solves
\[
b_u(\Delta u,v) + b''(\Delta u, \delta u, v_{R(u)}) = \ell(v) - b(u,v) = r(u,v),
\]
and aims to minimize the nonlinear dual residual instead of the linearized problem residual. 
}

\subsection{Area C}
\frame{
\frametitle{Proposed work: Area C}
\begin{columns}
\begin{column}{.48\textwidth}
\begin{itemize}
\item{\textbf{Completed: convection-dominated diffusion, Burgers, and a model problem for Navier-Stokes.}}
\item{\textbf{\textcolor{red}{Proposed}: Range of Reynolds/Mach numbers.}}
\item{\textbf{\textcolor{red}{Proposed}: Flow over a bump, Holden's ramp problem (shock-boundary layer interaction).}}
\item{\textbf{\textcolor{red}{Proposed (time permitting)}: airfoils, regularized Euler.}}
\end{itemize}
\end{column}
\begin{column}{.5\textwidth}
\begin{figure}
\centering
\includegraphics[scale = .275]{figs/ramp.png}
\end{figure}
\end{column}
\end{columns}
}

>>>>

\title[A DPG method for Compressible Flow]{A Discontinuous Petrov-Galerkin method for compressible flow problems}
\date{August 15, 2012}
\author[J. Chan, T. Ellis, L. Demkowicz, R. Moser]{Jesse Chan, Truman Ellis \\ Supervisors: Leszek Demkowicz, Robert Moser}

@@

}
\frame{
\frametitle{Compressible Navier-Stokes equations}
\begin{columns}[c]
\begin{column}{.475\textwidth}
Numerical difficulties:
\begin{itemize}
\item{} Resolving solution features (sharp, localized viscous-scale phenomena)
\begin{itemize}
\item{} Shocks
\item{} Boundary layers - resolution needed for drag/load
\item{} Turbulence (non-localized)
\end{itemize}
\item{} Nonlinear convergence and uniqueness of solutions
\item{} Stability of numerical schemes
\begin{itemize}
\item{} Coarse/adaptive grids
\item{} Higher order
\end{itemize}
\end{itemize}
\end{column}
\begin{column}{.475\textwidth}
\begin{figure}
\centering
\includegraphics[scale = .7]{figs/compFlowPics/airfoil_mesh.png}
\end{figure}
\end{column}
\end{columns}
\vspace{.2cm}
\begin{center}
Idea: begin first with the model problem of convection-diffusion.
\end{center}
}
\frame{
\frametitle{Robustness: convection-diffusion as a model problem}
\vspace{-.5cm}
\[
\div \left(\beta u\right) - \epsilon \Delta u = f, \quad \text{on }\Omega \in \mathbb{R}^3
\]
\vspace{-.5cm}
\begin{columns}[c]
\begin{column}{.49\textwidth}
In 1D: $\beta u' - \epsilon u'' = f$. Standard continuous Galerkin variational formulation: solve
\[
b(u,v) = \ell(v), \quad u,v\in H^1_0([0,1])
\]
where
\begin{align*}
b(u,v) &= \int_\Omega -\beta uv' + \epsilon u'v' \\
\ell(v) &= \int_\Omega f v 
\end{align*}
\end{column}
\begin{column}{.49\textwidth}
\begin{figure}
\centering
\includegraphics[scale=.35]{figs/GalerkinOscTight.png}
\caption{Solution for $f=1$. Oscillations in the standard Galerkin method for large Peclet numbers ${\rm Pe} \coloneqq \frac{h}{\epsilon}$.}
\end{figure}
\end{column}
\end{columns}
}

\frame{
\frametitle{Streamline-upwind Petrov-Galerkin (SUPG)}
SUPG solves $b_{\rm SUPG}(u,v) = l_{\rm SUPG}(v)$, where
\begin{align*}
b_{\rm SUPG}(u,v) &= b(u,v)+ \sum_{K} \int_{K} \tau \LRp{L_{\rm adv} v} L u \\
l_{\rm SUPG}(v) &= \ell(v) + \sum_{K} \int_{K} \tau \LRp{L_{\rm adv}v} f.
\end{align*}
\vspace{-.5cm}
\begin{columns}[c]
\begin{column}{.44\textwidth}
\vspace{-.5cm}
\begin{itemize}
\item{} $Lu = \div\left(\beta u\right) - \epsilon \triangle u$, $L_{\rm adv}u = \div \left(\beta u\right)$, and $\tau$ is a parameter. 
%\item{} Recovers ``exact'' artificial diffusion for $f=0$.
\item{} Effective for $f\neq 0$. %``exact'' artificial diffusion.
\item{} \emph{Residual-based} stabilization. 
\end{itemize}
\end{column}
\begin{column}{.54\textwidth}
\begin{figure}
\centering
\includegraphics[scale=.38]{figs/SUPG.png}
%\caption{Exact, Bubnov-Galerkin, and SUPG solutions.}
\end{figure}
\end{column}
\end{columns}
}

@@

%\frame{
%\frametitle{Regularization}
%Discontinuous inflow conditions on $\widehat{f_n}$, out of alignment with mesh. Boundary condition $\epsilon \grad u \cdot n = 0$ at outflow.
%\begin{figure}
%\centering
%\subfigure{
%\includegraphics[scale=.33]{figs/discontinuous.png}
%}
%\subfigure{
%\includegraphics[scale=.305]{figs/discontinuousL2rates.png}
%}
%\caption{Discontinuous hat inflow data with $\epsilon = 1e-5$, convection solution errors.}
%\end{figure}
%}
\frame{
\frametitle{Local conservation}
Given stiffness matrix $K$, can introduce Lagrange multipliers to enforce element-wise constraints, leading to the saddle point problem
\[
\arr{K}{B}{B^T}{0}\vecttwo{u}{\lambda} = \vecttwo{f}{c}.
\]
where the Lagrange multipliers $\lambda$ can be locally condensed. We use the test norm
\begin{align*}
\|\left(v,\tau\right)\|_{V,K}^2 = \left|\int_K v\right|^2 + \epsilon \|\grad v\|^2 + \|\beta \cdot \grad v\|^2 +&\\
\| \div \tau\|^2 + \min\left\{\frac{1}{\epsilon},\frac{1}{|K|}\right\}\|\tau\|^2&.
\end{align*}
}

\frame{
\frametitle{Additional examples}
\begin{figure}
\centering
\subfigure[Advection skew]{
\includegraphics[scale = .175]{figs/Conservative/hughes.png}
}
\subfigure[Flow over a step]{
\includegraphics[scale = .175]{figs/Conservative/step.png}
}
\subfigure[Double glazing]{
\includegraphics[scale = .182]{figs/Conservative/doubleGlazing.png}
}
\end{figure}
Locally conservative version of DPG performs very similarly to the standard DPG in all cases.
}

@@

\begin{figure}
\centering
\subfigure{
\includegraphics[scale=.21]{figs/vortex.png}
}
\subfigure{
\includegraphics[scale=.2015]{figs/vortexMesh.png}
}
\caption{Steady vortex problem with $\epsilon = 1e-4$.}
\end{figure}
\vspace{-.25cm}
}

\frame{
\frametitle{Current work - Hemker problem}

\begin{figure}
\centering
\subfigure{
\includegraphics[scale = .225]{figs/Conservative/Hemker/WithoutMesh.png}
}
\subfigure{
\includegraphics[scale = .225]{figs/Conservative/Hemker/WithMesh.png}
}
\caption{Convection-diffusion solution to the Hemker problem.}
\end{figure}
}
\frame{
\frametitle{Current work - high-lift airfoil}
\begin{figure}
\centering
\subfigure{
\includegraphics[scale = .1875]{figs/Conservative/Airfoil-Conservative-1E-3/solution.png}
}
\subfigure{
\includegraphics[scale = .175]{figs/Conservative/Airfoil-Conservative-1E-3/mesh7.png}
}
\caption{Convection-diffusion solution over the high-lift airfoil, with advection at an angle, $\epsilon = 1e-3$, with $p=3$. Mesh is generated by \emph{Triangle}.}
\end{figure}
}

@@

\section{Future work}
%\subsection{Area A}
%\frame{
%\frametitle{Proposed work: Area A}
%\begin{itemize}
%\item{\textbf{Completed: Prove robustness of DPG method for the scalar convection-diffusion problem.}}
%
%We have introduced a test norm under which the DPG method robustly bounds the $L^2$ error in the field variables $u$ and the scaled stress $\sigma$. Numerical results confirm the theoretical bounds given. 
%
%\item{\textbf{\textcolor{red}{Proposed}: Attempt analysis of the linearized Navier-Stokes system.}}
%
%We hope to analyze the linearized Navier-Stokes equations to determine an optimal extrapolation of the test norm for the scalar convection-diffusion problem to systems. 
%
%\end{itemize}
%}

%\subsection{Area B}
%
\frame{
%\frametitle{Proposed work: Area B}
%\item{\textbf{Completed: Collaborative work with Nathan Roberts on the higher order parallel adaptive DPG code Camellia.}}
%
%Numerical experiments use the higher-order adaptive codebase Camellia, built by Nathan Roberts upon the Trilinos library. %The framework for arbitrary-irregularity anisotropic refinements in both $h$ and $p$ is in place, and the code is partially parallelized.

\frametitle{Future work: anisotropic refinements and $hp$-schemes.}

The error representation function drives refinement effectively, and we hope to generalize its use to anisotropic adaptive schemes. 
\begin{figure}
\centering
\subfigure{\includegraphics[scale = .175]{figs/anisotropy/b5.png}}
\subfigure{\includegraphics[scale = .18]{figs/anisotropy/v6.png}}
\caption{Anisotropic mesh for a convection-diffusion boundary layer.}
\end{figure}
}

\frame{
\frametitle{Future work: distributed iterative static condensation.}
\[
Ku = \arr{A}{B}{B^T}{D}\vecttwo{u_{\rm flux}}{u_{\rm field}} = \vecttwo{f}{g} = l
\]
where $D$ has a block-diagonal structure. The system can be reduced to yield the condensed system
\[
\left(A-B D^{-1} B^T\right)u_{\rm flux} = f - B D^{-1} g
\]
where $D^{-1}$ can be inverted block-wise. For FE stiffness matrices under the Laplace equation, the Schur complement has reduced condition number of $O(h^{-1})$ as opposed to $O(h^{-2})$.\footnote{\bibentry{schurComplement}}
}

\frame{  
\frametitle{Future work: a Nonlinear Hessian-based DPG method.}
Given a nonlinear variational problem $b(u,v) = \ell(v)$, linear in $v$ but not in $u$, beginning with the \textit{nonlinear} dual residual 
\[
J(u_h) = \frac{1}{2}\|B(u_h)-\ell\|_{V'}^2 \coloneqq\frac{1}{2} \sup_{v\in V\setminus\{0\}} \frac{| b(u_h,v)-\ell(v)|^2}{\nor{v}_V^2}.
\]
produces a Hessian-based DPG method, which solves
\[
b_u(\Delta u,v) + b''(\Delta u, \delta u, v_{R(u)}) = \ell(v) - b(u,v) = r(u,v),
\]
and aims to minimize the nonlinear dual residual instead of the linearized problem residual. 
}

\frame{
\frametitle{Future work: applications}
\begin{columns}
\begin{column}{.48\textwidth}
\begin{itemize}
\item{{Solutions for all problems over a range of Reynolds/Mach numbers.}}
\item{{Flow over a bump, Holden's ramp problem (shock-boundary layer interaction).}}
\item{{Time permitting: airfoils, regularized Euler.}}
\end{itemize}
\end{column}
\begin{column}{.5\textwidth}
\begin{figure}
\centering
\includegraphics[scale = .275]{figs/ramp.png}
\end{figure}
\end{column}
\end{columns}
}
